---
title: "Variation in Penguin Body Mass Across Islands in the Palmer Archipelago"
author: "Tyler Rayman"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    fig_caption: true
bibliography: references.bib
csl: apa.csl
header-includes:
  - \usepackage{caption}
  - \captionsetup{justification=raggedright,singlelinecheck=false,labelfont=bf}
---

# Abstract
Body mass is an important biological trait that can vary across species populations due to ecological and environmental factors. In penguins specifically, body mass differences may reflect variation in habitat conditions, resource availability, or individual lifespan among locations. This study examined whether penguin body mass differs across three islands in the Palmer Archipelago using the Palmer Penguins dataset. Body mass measurements from penguins sampled on the islands of Biscoe, Dream, and Torgersen were analyzed using a one-way analysis of variance (ANOVA). Mean body mass differed significantly among islands, and therefore a Tukey's HSD test was applied to locate the differences in means.

# Introduction

Body size is an important biological trait that ... 
In penguins, body mass ...  

This study explores whether penguin body mass differs across islands using the Palmer Penguins dataset (Horst et al., 2020).

# Methods

## Data Source

Data were obtained from the `palmerpenguins` R package (Horst et al., 2020).  
The raw dataset was exported to CSV using `scripts/01_load_raw_penguins.R`.  

Data cleaning was performed using `scripts/02_clean_penguins.R`, which:
- Selected the variables `island` and `body_mass_g`
- Removed rows containing missing values
- Generated a processed dataset for analysis

## Statistical Analysis

All analyses were performed in R.

# Results
```{r load-data, echo=FALSE}
penguins <- read.csv("../data/data_processed/penguins_body_mass_by_island.csv")
```
Observations across penguins measured on the three islands are shown below in a summary table. 
```{r summary-stats, echo=FALSE}

means <- tapply(penguins$body_mass_g, penguins$island, mean)
sds <- tapply(penguins$body_mass_g, penguins$island, sd)
ns <- tapply(penguins$body_mass_g, penguins$island, length)

summary_stats <- data.frame(
  Island = names(means),
  Mean = round(means, 1),
  SD = round(sds, 1),
  N = ns
)

# save the table to the outputs folder
write.csv(summary_stats,
          "../outputs/penguin_summary_statistics.csv",
          row.names = FALSE)

# display the table in the manuscript
knitr::kable(summary_stats,
             caption = "Summary statistics for penguin body mass across islands.",
             row.names = FALSE)

```
To visualize the distribution of body mass across islands, a boxplot was generated (Figure 1). Penguins sampled from Biscoe Island appear to have higher body mass values overall compared with those from Dream and Torgersen islands.
```{r body-mass-figure, echo=FALSE, fig.cap="Distribution of penguin body mass across the three islands.", fig.align="center"}
boxplot(body_mass_g ~ island,
        data = penguins,
        xlab = "Island",
        ylab = "Body Mass (g)",
        col = "lightblue")
```
A one-way analysis of variance (ANOVA) revealed a significant effect of island on body mass (F(2, 339) = 110, p < 0.001). Post-hoc comparisons using Tukey’s Honestly Significant Difference (HSD) test indicated that penguins from Biscoe Island had significantly greater body mass than those from Dream and Torgersen islands (p < 0.001 for both comparisons), while body mass did not differ significantly between Dream and Torgersen islands (p = 0.998).
```{r anova-test, echo=FALSE}
anova_model <- aov(body_mass_g ~ island, data = penguins)
anova_result <- summary(anova_model)

anova_table <- anova_result[[1]]
rownames(anova_table)[1] <- "Island"

knitr::kable(anova_table, digits = 2)
```

```{r tukey-test, echo=FALSE, include=FALSE}
tukey_result <- TukeyHSD(anova_model)
```

Mean penguin body mass varied across islands (Figure 1). A one-way analysis of variance (ANOVA) revealed a significant effect of island on body mass (F(2, 339) = 110, p < 0.001). Post-hoc comparisons using Tukey’s Honestly Significant Difference (HSD) test indicated that penguins from Biscoe Island had significantly greater body mass than those from Dream and Torgersen islands (p < 0.001 for both comparisons), while body mass did not differ significantly between Dream and Torgersen islands (p = 0.998).

# Discussion

Differences in body mass across islands may reflect a multitude of reasons. Based on the ANOVA test results, penguins measured on the island of Biscoe have significantly larger mass than those on the other two islands.

Because this dataset is observational, results should not be interpreted as causal.

# Data and Code Availability
All data processing and analysis scripts are available in the scripts directory. 

# References

Blackburn, T. M., Gaston, K. J., & Loder, N. (1999). Geographic gradients in body size: A clarification of Bergmann’s rule. *Diversity and Distributions, 5*(4), 165–174. https://doi.org/10.1046/j.1472-4642.1999.00046.x

Gorman, K. B., Williams, T. D., & Fraser, W. R. (2014). Ecological Sexual Dimorphism and Environmental Variability within a Community of Antarctic Penguins (Genus Pygoscelis). *PLoS ONE, 9*(3), e90081. https://doi.org/10.1371/journal.pone.0090081

M. Horst, A., Presmanes Hill, A., & B. Gorman, K. (2022). Palmer Archipelago Penguins Data in the palmerpenguins R Package—An Alternative to Anderson’s Irises. *The R Journal, 14*(1), 244–254. https://doi.org/10.32614/RJ-2022-020

