---
title: "Variation in Penguin Body Mass Across Islands in the Palmer Archipelago"
author: "Tyler Rayman"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    number_sections: false
bibliography: references.bib
csl: apa.csl
---

# Abstract

# Introduction

Body size is an important biological trait that ... 
In penguins, body mass ...  

This study explores whether penguin body mass differs across islands using the Palmer Penguins dataset (Horst et al., 2020).

# Methods

## Data Source

Data were obtained from the `palmerpenguins` R package (Horst et al., 2020).  
The raw dataset was exported to CSV using `scripts/01_load_raw_penguins.R`.  

Data cleaning was performed using `scripts/02_clean_penguins.R`, which:
- Selected the variables `island` and `body_mass_g`
- Removed rows containing missing values
- Generated a processed dataset for analysis

## Statistical Analysis

All analyses were performed in R.

# Results

Mean body mass varied across islands (Figure 1).
```{r load-data, echo=FALSE}
penguins <- read.csv("../data/data_processed/penguins_body_mass_by_island.csv")
```

```{r summary-stats, echo=FALSE}
summary_stats <- data.frame(
  island = names(tapply(penguins$body_mass_g, penguins$island, mean)),
  mean = tapply(penguins$body_mass_g, penguins$island, mean),
  sd = tapply(penguins$body_mass_g, penguins$island, sd),
  n = tapply(penguins$body_mass_g, penguins$island, length)
)

knitr::kable(summary_stats, digits = 2)
```

```{r body-mass-figure, echo=FALSE, fig.cap="Distribution of penguin body mass across the three islands.", fig.path="../outputs/"}
boxplot(body_mass_g ~ island,
        data = penguins,
        xlab = "Island",
        ylab = "Body Mass (g)",
        col = "lightblue")
```
A one-way analysis of variance (ANOVA) revealed a significant effect of island on body mass (F(2, 339) = 110, p < 0.001). Post-hoc comparisons using Tukey’s Honestly Significant Difference (HSD) test indicated that penguins from Biscoe Island had significantly greater body mass than those from Dream and Torgersen islands (p < 0.001 for both comparisons), while body mass did not differ significantly between Dream and Torgersen islands (p = 0.998).
```{r anova-test, echo=FALSE}
anova_model <- aov(body_mass_g ~ island, data = penguins)
anova_result <- summary(anova_model)

knitr::kable(anova_result[[1]], digits = 2)
```

```{r tukey-test, echo=FALSE, include=FALSE}
tukey_result <- TukeyHSD(anova_model)
```

Mean penguin body mass varied across islands (Figure 1). A one-way analysis of variance (ANOVA) revealed a significant effect of island on body mass (F(2, 339) = 110, p < 0.001). Post-hoc comparisons using Tukey’s Honestly Significant Difference (HSD) test indicated that penguins from Biscoe Island had significantly greater body mass than those from Dream and Torgersen islands (p < 0.001 for both comparisons), while body mass did not differ significantly between Dream and Torgersen islands (p = 0.998).


# Discussion

Differences in body mass across islands may reflect... insert reasons. 

Because this dataset is observational, results should not be interpreted as causal.

# Reproducibility

All data processing and analysis scripts are available in the `scripts` directory.  

# References

